---
title: "Basic MIDI mangling functions"
author: "Matt Crump"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Basic MIDI mangling functions}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(midiblender)
```

A collection of tools for mangling basic aspects of midi files.

## mangle_positive_timing()

Add/subtract random ticks to the time column from a uniform distribution. Set the range of the distribution with `ticks_amount` to control the amount of jitter. Any resulting time values that are negative are set to zero. This function does not add jitter to spread out simultaneous messages, so chords do not become jittered triads.

Useful for subtle humanization, or to absolutely wreck the timing information.

Notes: This function was written to work in a dplyr pipe. It might be nice to hide that part inside the function so this can be run directly on the df...not sure yet, maybe both.

```{r, eval=FALSE}
library(dplyr)

# import midi
mario <- midi_to_object("all_overworld.mid")
list2env(mario, .GlobalEnv)

# a dplyr pipe to apply jitter non-meta messages
midi_df %>%
  mutate(time = case_when(
    meta == FALSE ~ mangle_positive_timing(time, ticks_amount = 5),
    meta == TRUE ~ time
  )) -> copy_df

# update miditapyr df
miditapyr_object$midi_frame_unnested$update_unnested_mf(copy_df)

#write midi file to disk
miditapyr_object$write_file("jitter_mario.mid")

```

