---
title: "Midi data frame constructors with dplyr"
author: "Matt Crump"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Midi data frame constructors with dplyr}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(midiblender)
```

The midi-construct functions are helpers for building a valid `midi_df` data frame using `dplyr` syntax.

These functions can be useful for constructing blocks of meta_messages, or whole midi_dfs. The functions are mostly wrapping `dplyr::add_row()`. The advantage of wrapping is that important parameters become available with auto-complete.

```{r, eval=FALSE}
new_midi_df <- create_empty_midi_df() %>% # initialize 
  add_meta_track_name(name = "My track") %>%
  add_meta_tempo(tempo = 500000) %>%
  add_meta_time_sig(numerator = 4,
                    denominator = 4,
                    clocks_per_click = 36,
                    notated_32nd_notes_per_beat = 8) %>%
  add_program_change(program = 0,
                     channel = 0) %>%
  add_control_change(control = 0,value = 0) %>%
  add_note_on(time = 0,note = 60) %>%
  add_note_off(time = 8,note = 60) %>%
  add_meta_end_of_track()

knitr::kable(new_midi_df)

##############
# write midi

#Initialize new pyramidi object
new_pyramidi_object <- pyramidi::MidiFramer$new()
# update ticks per beat
new_pyramidi_object$ticks_per_beat <- 96L
# update object with new midi df
new_pyramidi_object$mf$midi_frame_unnested$update_unnested_mf(new_midi_df)
# write to midi file
new_pyramidi_object$mf$write_file("file_name.mid")
  
```

## Mono lead example

This code chunk generates a random series of notes from a vector of possible notes.

```{r, eval=FALSE}
# note parameters
num_notes <- 128
note_duration <- 24
possible_notes <- c(60, 63, 65, 66, 67, 70, 72, 75)

new_midi_df <- create_empty_midi_df() %>% # initialize 
  add_meta_track_name(name = "My track") %>%
  add_meta_tempo(tempo = 500000) %>%
  add_meta_time_sig(numerator = 4,
                    denominator = 4,
                    clocks_per_click = 36,
                    notated_32nd_notes_per_beat = 8) %>%
  add_program_change(program = 0,
                     channel = 0) %>%
  add_control_change(control = 0,value = 0) %>%
  #use dplyr::add_row to add a bunch of notes
  add_row(
    i_track = 0,
    meta = FALSE,
    type = rep(c("note_on", "note_off"),num_notes),
    time = rep(c(0, note_duration), num_notes),
    note = rep(sample(possible_notes,
                  size = num_notes,
                  replace= TRUE),each=2)
  ) %>%
  add_meta_end_of_track()


##############
# write midi

#Initialize new pyramidi object
new_pyramidi_object <- pyramidi::MidiFramer$new()
# update ticks per beat
new_pyramidi_object$ticks_per_beat <- 96L
# update object with new midi df
new_pyramidi_object$mf$midi_frame_unnested$update_unnested_mf(new_midi_df)
# write to midi file
new_pyramidi_object$mf$write_file("file_name.mid")

```


